---
title: "report_kiki"
author: "kiki"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
We start by analyzing some models where the response variable "doctorco" is transformed into the binary response variable "ifvisit". This binary variable is equal to 0 if "doctorco" is 0, and is equal to 1 otherwise. By doing so, we are modelling the number of people that went to a doctor's visit at least once in the last two weeks.
To start we import the data set and create the "ifvisit" variable. From the following graph we can see the skewness of the data set regarding this variable:

```{r echo=FALSE, warning=FALSE, include=TRUE}
load("/home/sebas/Documents/chris/statistica/FINAL_PROJECT/HealthCareAustralia.rda")
data = ex3.health
data$ifvisit = ifelse(data$doctorco == 0, 0, 1)
set.seed(42)
train_indices <- sample(seq_len(nrow(data)), 0.8 * nrow(data))
train_data <- data[train_indices, ]
test_data <- data[-train_indices, ]
hist(data$ifvisit, main = "Original data set")
```

We can try fitting a GAM using the obtained data set but the results are poor (we get a maximum of around 19% explained deviance):
```{r echo=TRUE, warning=FALSE, message=FALSE}
library("mgcv")
model_gam <- gam(ifvisit ~ s(hospdays) + s(actdays) + age*prescrib + freepoor + hscore + medicine + illness, data = train_data, family = binomial(link = "logit"))
summary(model_gam)
```
```{r echo=FALSE, warning=FALSE, message=FALSE}
print(paste("AIC (GAM):", round(AIC(model_gam),2)))
predicted_counts <- round(predict(model_gam, newdata = test_data, type = "response"))
true_counts <- test_data$ifvisit
rmse <- mean(abs(predicted_counts - true_counts))
print(paste("MAE (GAM):", round(rmse,4)))
print(paste("Number of true ifvisit:",sum(true_counts)))
print(paste("Number of ifvisit predicted:", sum(predicted_counts)))
par(mfrow = c(1, 2))
hist(true_counts,ylim = c(0, 1000), main = "True ifvisit")
hist(predicted_counts,ylim = c(0, 1000), main = "Predicted ifvisit")
par(mfrow = c(1, 1))
```

Since we transformed the problem into a binary classification problem we can use techniques to balance the data set. One option is to use ROSE (Random Over-Sampling Examples), which is a method used for oversampling the minority class in binary classification problems to balance the data set. It involves generating synthetic examples from the existing minority class instances. This can be achieved by randomly selecting a minority class instance and introducing variations to create new synthetic instances.
```{r warning=FALSE, echo=TRUE, message=FALSE}
library(ROSE)
data.rose <- ROSE(ifvisit ~ ., data = data, seed = 1, hmult.majo = 0)$data
hist(data.rose$ifvisit, main = "Balanced data set")
```
```{r warning=FALSE, echo=FALSE, message=FALSE}
set.seed(42)
train_indices <- sample(seq_len(nrow(data.rose)), 0.8 * nrow(data.rose))
train_data <- data.rose[train_indices, ]
test_data <- data.rose[-train_indices, ]
```

We can try again fitting a GAM on this new balanced data set and see the results:
```{r warning=FALSE, echo=FALSE, message=FALSE}
library("mgcv")
model_gam <- gam(ifvisit ~ s(actdays) + s(hospadmi) + s(nonpresc), data = train_data, family = binomial(link = "logit"))
summary(model_gam)
```
```{r echo=FALSE, warning=FALSE, message=FALSE}
print(paste("AIC (GAM):", round(AIC(model_gam),2)))
predicted_counts <- round(predict(model_gam, newdata = test_data, type = "response"))
true_counts <- test_data$ifvisit
rmse <- mean(abs(predicted_counts - true_counts))
print(paste("MAE (GAM):", round(rmse,5)))
print(paste("Number of true ifvisit:",sum(true_counts)))
print(paste("Number of ifvisit predicted:", sum(predicted_counts)))
par(mfrow = c(1, 2))
hist(true_counts,ylim = c(0, 1000), main = "True ifvisit")
hist(predicted_counts,ylim = c(0, 1000), main = "Predicted ifvisit")
par(mfrow = c(1, 1))
```


```{r echo=FALSE, warning=FALSE, message=FALSE}

```

